---
title: "Data Science Challenge 3 red foxes"
author: Louise Litrico (s1814797@ed.ac.uk)
date: 10/Nov/2021
output: 
  html_document:
    toc: true
    toc_float: true
---

## Introduction

The data used in this study was collected by the *Living Planet Index* (LPI 2016. [Living Planet Index database](https://livingplanetindex.org/home/index). 2016., available to use for scientific analysis or research and conservation purposes. 

# Data manipulation 
```{r, warning=FALSE}
library(tidyverse)
```

```{r}
LPI_species <- load("LPI_species.Rdata")
```

I chose to study the red fox populations and their changes through time.
```{r}
fox_data <- LPI_species %>% 
  filter(Common.Name == "Red fox") %>% 
  mutate(year = as.character(.$year)) %>% 
  mutate(year = parse_number(.$year)) %>%  # need year as numerical values
  mutate(origin = case_when(grepl("Yes", .$Native) ~ "Native", 
                            grepl("Yes", .$Invasive) ~ "Invasive")) %>% 
  select(-Native, -Invasive, -Alien) %>%   # reduce to 1  column for origin instead of 3 
  group_by(Data.source.citation) %>% 
  mutate(lengthyear = (max(year) - min(year))) %>%  # look at length of studies
  ungroup()
```

# General data information
* the data range from 1970 up to 2006
* the data come from 10 studies, carried out in 24 location found in 5 different countries located in 3 regions of the world
* each study uses a different sampling method to get an estimate of the population of red foxes in their are
* although there are 10 studies and 10 sampling methods, the dataset includes 11 units

```{r}

```

Although the high number of divergent transitions after warmup can create doubt concerning the quality of the predictions, the model converged well and the model fits the data nicely as well, showing that it can still be used to explain and predict red fox population changes. 
